Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-15 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: parspec section; reinstated mathcal L for likelihood

We start by reviewing the parametrized ringdown spin expansion coefficients
formalism (ParSpec) introduced in Ref.~\cite{Maselli:2019mjd}, but following
some of the notation of Ref.~\cite{Carullo:2021dui}.
%
A general procedure to describe deviations from the Kerr quasinormal mode frequencies $\omega_{J}^{\rm K}$ and
damping time $\tau_{J}^{\rm K}$ is to write them as,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{J} &= \omega^{\rm K}_{J} \, (1 + \delta \omega_{J}), \\
\tau_{J}   &= \tau^{\rm K}_{J}   \, (1 + \delta \tau_{J}),
\end{align}
\end{subequations}
%
as done e.g., in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Carullo:2018sfu}.
%
Here we use $J$ to collectively represents the QNM labels $\{l,\, m,\, n\}$,
where $l$ and $m$ are the multipole indices, and $n$ the overtone number.
%
We work exclusively with $\ell = m = 2$ and $n = 0$ (fundamental mode) in this work.

We expand the Kerr expression as:
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{J} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{J} \, \left( 1 + \gamma \delta \omega^{(n)}_{J} \right), \\
\tau_{J}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{J}   \, \left( 1 + \gamma \delta \tau^{(n)}_{J} \right),
\end{align}
\end{subequations}
%
where $M$ and $\chi$ are the detector-frame mass and dimensionless spin of the
black hole; $\omega_{K}^{n}$, $\tau_{K}^{n}$ are dimensionless coefficients
of the spin expansion for Kerr QNMs; $\delta\omega_{K}^{n}$, $\delta\tau_{K}^{n}$ are
source-independent dimensionless coefficients that characterize the corrections to the Kerr QNM at each spin-order;
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on the beyond-GR modification is encapsulated in the dimensionless
parameter $\gamma$ given by,
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
included to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

In principle, a modification to GR would also affect $M$ and $\chi$ and the expansion should be written
in terms of non-GR mass and spin, say $\bar{M}$ and $\bar{\chi}$. If we assume that the beyond-GR corrections
are included perturbatively, the modifications to the BH mass and spin can be absorved into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their corresponding GR values.
%

\hs{TODO: in figures and tables use $N_{\rm max} = 0$ ($N_{\rm max} = 1$) instead of $j=0$ ($j=0,1$).}
\hs{TODO: say that we check a posteriori if our PE runs agree with this.}

@@

Let us now establish the connection between theory-agnostic framework of the
$\pSEOB$ waveform model and the theory-specific QNM calculations.
%
We begin by expanding $\omega$ and $\tau$, given by Eqs.~\eqref{eq:kerr_expansion} as,
%
\begin{subequations}
\begin{align}
M \omega_{220} &= \gamma \left( \delta\omega^{(0)} \omega^{(0)} + \chi \delta\omega^{(1)} \omega^{(1)} \right)
+ \sum_{j=0}^{N} \chi^{j} \omega^{j}\,,
\\
\tau_{220}/M   &= \gamma \left( \delta\tau^{(0)} \tau^{(0)} + \chi \delta\tau^{(1)} \tau^{(1)} \right)
+ \sum_{j=0}^{N} \chi^{j} \tau^{j}\,,
\end{align}
\label{eq:qnm_nongr_iso}
\end{subequations}
%
where we are already focusing on the dominant $\ell = m = 2$, $n=0$ QNM and
pulled out of the sum all beyond-GR corrections, rescriting ourselves to the nonspinning ($j=0$) and linear-order in spin ($j=1$)
corrections to the Kerr QNMs, encoded in the coefficients $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$.

@@

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) \, P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is the union between the GR waveform
model parameters~$\bm{\theta}$, detailed in Sec.~\ref{sec:review_pSEOB}, and
the non-GR parameter of $\ell$, i.e.,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\label{eq:def_params}
\end{equation}
%
which, we recall, set the characteristic lengthscale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.

>>>>

We start by reviewing the parametrized ringdown spin expansion coefficients (ParSpec)
framework introduced in Ref.~\cite{Maselli:2019mjd}, but following
some of the notation used in Ref.~\cite{Carullo:2021dui}.

A general procedure to describe deviations from the Kerr QNM frequencies $\omega_{J}^{\rm K}$ and
damping times $\tau_{J}^{\rm K}$ is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{J} &= \omega^{\rm K}_{J} \, (1 + \delta \omega_{J}), \\
\tau_{J}   &= \tau^{\rm K}_{J}   \, (1 + \delta \tau_{J}),
\end{align}
\end{subequations}
%
where $\delta\omega_{J}$ and $\delta\tau_{J}$ are the deformation parameters.
%
Here we use the subscript $J$ to represent the QNM labels $\{l,\, m,\, n\}$,
where $l$ and $m$ are the multipole indices, and $n$ the overtone number.
%
We will work exclusively with the $\ell = m = 2$ and $n = 0$ (fundamental mode)
QNM in this work, that is, $J = \{2,\, 2,\, 0\}$.
%
This approach was adopted, e.g., in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Carullo:2018sfu}.

A caveat of the parametrizations of the form~\eqref{eq:general_deviation}, is that
$\delta\omega_{J}$ and $\delta\tau_{J}$ depend, in general, on the BH's source mass and spin.
%
Ideally, one would like to explicitly reinstate the dependence the source-frame
BH mass and spin for it would
%
%
(i) introduce deformation parameters which can be determined, once and for
all, from a specific gravity theory, and
%
(ii) make it easy to combine constraints from multiple source.

In the ParSpec framework this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame mass and
spin, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{J} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{J} \, \left( 1 + \gamma \delta \omega^{(n)}_{J} \right), \\
\tau_{J}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{J}   \, \left( 1 + \gamma \delta \tau^{(n)}_{J} \right),
\end{align}
\end{subequations}
%
where $M$ and $\chi$ are the detector-frame mass and dimensionless spin of the
black hole; $\omega_{K}^{n}$, $\tau_{K}^{n}$ are dimensionless coefficients
of the spin expansion for Kerr QNMs; $\delta\omega_{K}^{n}$, $\delta\tau_{K}^{n}$ are
source-independent dimensionless coefficients that characterize the corrections to the Kerr QNM at each spin-order;
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on the beyond-GR modification is encapsulated in the dimensionless
parameter $\gamma$ given by,
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
included to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

In principle, a modification to GR would also affect $M$ and $\chi$ and the expansion should be written
in terms of non-GR mass and spin, say $\bar{M}$ and $\bar{\chi}$. If we assume that the beyond-GR corrections
are included perturbatively, the modifications to the BH mass and spin can be absorved into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.

% \hs{TODO: in figures and tables use $N_{\rm max} = 0$ ($N_{\rm max} = 1$) instead of $j=0$ ($j=0,1$).}
\hs{TODO: say that we check a posteriori if our PE runs agree with this.}

@@

Let us now establish the connection between theory-agnostic framework of the
$\pSEOB$ waveform model and the theory-specific QNM calculations.
%
We begin by expanding $\omega$ and $\tau$, given by Eqs.~\eqref{eq:kerr_expansion} as,
%
\begin{subequations}
\label{eq:qnm_nongr_iso}
\begin{align}
M \omega_{220} &= \gamma \left( \delta\omega^{(0)} \omega^{(0)} + \chi \delta\omega^{(1)} \omega^{(1)} \right)
+ \sum_{n=0}^{N_{\rm max}} \chi^{j} \omega^{j}\,,
\\
\tau_{220}/M   &= \gamma \left( \delta\tau^{(0)} \tau^{(0)} + \chi \delta\tau^{(1)} \tau^{(1)} \right)
+ \sum_{n=0}^{N_{\rm max}} \chi^{j} \tau^{j}\,,
\end{align}
\end{subequations}
%
where, recall, we focus on the $\ell = m = 2$, $n=0$ QNM and have pulled out of
the sum all beyond-GR corrections, restricting ourselves to the nonspinning
($n=0$) and linear-order in spin ($n=1$) corrections to the Kerr QNMs.
%
These corrections are encoded in the coefficients $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$.

@@

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) \, \mathcal{L}(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$\mathcal{L}(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is the union between the GR waveform
model parameters~$\bm{\theta}$, detailed in Sec.~\ref{sec:review_pSEOB}, and
the non-GR parameter of $\ell$, i.e.,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\label{eq:def_params}
\end{equation}
%
which, we recall, set the characteristic lengthscale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}.
